\Large{Adversarial Training against Systematic Uncertainty}

### 2016-10-24 tex/paper.tex
physics example

In this work, we repurpose adversarial training as a means to constraint the
predictive model $f$ in order to satisfy Eqn.~\ref{eqn:criterion}. As
illustrated in Figure~\ref{fig:architecture}, we pit $f$ against an adversary
model $r := p_{\theta_r}(z | f(X;\theta_f)=s)$ of parameters $\theta_r$ and
associated loss ${\cal L}_r(\theta_f, \theta_r)$. This model takes  as input
realizations $s$ of $f(X; \theta_f)$, for the current value $\theta_f$ of $f$
parameters, and produces as output a function $p_{\theta_r}(z | f(X;\theta_f)=s)$
modeling the posterior probability density that $z$ parameterizes the sample
observed as $s$.
Intuitively, if $p(f(X; \theta_f)=s|z)$ varies with $z$,
then the corresponding correlation can be captured by $r$. By contrast, if
$p(f(X; \theta_f)=s|z)$ is invariant with $z$, as we require, then $r$ should
perform poorly and be close to random guessing. Training $f$ such that it
additionally minimizes the performance of $r$ therefore acts as a regularization
towards Eqn.~\ref{eqn:criterion}.

@@

On many problems of interest, let us also note that such a classifier may
not exist because the nuisance parameter directly shapes the decision boundary,
in which cases the lower bound $ H({Y|X}) - H(Z) < {\cal L}_f(\theta_f) - H({Z|f(X;\theta_f)})$ is strict: $f$ can either be an optimal classifier or a
pivotal quantity, but not both simultaneously. In this situation, it is natural
to rewrite the value function $E$  as
\begin{equation}
    E_\lambda(\theta_f, \theta_r) = {\cal L}_f(\theta_f) - \lambda {\cal L}_r(\theta_f, \theta_r),
\end{equation}
where $\lambda \geq 0$ is a hyper-parameter controlling the trade-off between
the performance of $f$ and its independence with respect to the nuisance
parameter. Setting $\lambda$ to a large value will preferably enforces $f$ to
be pivotal while setting $\lambda$ close to $0$ will rather constraint $f$ to be
optimal.

@@

Assuming a gaussian prior $z \sim {\cal N}(0,1)$ of unit variance, we start by
generating training data $\{ x_i, y_i, z_i \}_{i=1}^N$, from which we build a
neural network classifier $f$ minimizing ${\cal L}_f(\theta_f)$ without
adversarial training. The network architecture comprises 2 dense hidden layers of 20
nodes with ReLU activations, followed  by a dense output layer with a single
node with a sigmoid activation. As shown in the upper plots of
Figure~\ref{fig:toy}, the resulting classifier is not pivotal, as the
conditional probability densities of its decision scores $f(X;\theta_f)$ show
large discrepancies between values $z$ of the nuisance. While not shown here, a
classifier trained only from data generated at the nominal value $Z=0$ would
also not be pivotal.

Let us now consider the adversarial training of $f$ against an adversary $r$
implemented as a mixture density network modeling $Z|f(X;\theta_f)$ as a mixture
of five gaussians. As for $f$, the network architecture comprises 2 dense hidden
layers of 20 nodes with ReLU activations, but is followed by an output layer of
15 nodes corresponding to the means, standard deviations and mixture
coefficients of the five gaussians. Output nodes for the mean values come with
linear activations, nodes for the standard deviations with exponential
activations to ensure positivity, while nodes for the mixture coefficients
implement the softmax function to ensure positivity and normalization. When
running Algorithm~\ref{alg:adversarial-training} as initialized with the
classifier $f$ obtained previously, adversarial training effectively reshapes
the decision function so it that becomes almost independent on the nuisance
parameter, as shown in the lower plots of Figure~\ref{fig:toy}. In particular,
the conditional probability densities of the decision scores $f(X;\theta_f)$ are
now very similar to each other, indicating only a small residual  dependency on the
nuisance, as theoretically expected. The dynamics of adversarial training is
illustrated in Figure~\ref{fig:toy-training}, where the losses ${\cal L}_f$,
${\cal L}_r$ and ${\cal L}_f - \lambda {\cal L}_r$ are evaluated after each
iteration of Algorithm~\ref{alg:adversarial-training}. In the first iterations,
we observe that the global objective ${\cal L}_f - \lambda {\cal L}_r$ is
minimized by making the classifier less accurate, hence the corresponding
increase of ${\cal L}_f$, but which results in a classifier that is more
pivotal, hence the corresponding increase of ${\cal L}_r$ and the total net
benefit. As learning goes, minimizing $E$ then requires making predictions that
are more accurate, hence decreasing ${\cal L}_f$, or that are even less dependent on
$Z$, hence shaping $p_{\theta_r}$ towards the prior $p(z)$. Indeed, ${\cal L}_f$
eventually starts to slightly decrease, while remaining lower bounded by
$\min_{\theta_f} {\cal L}_f(\theta_f)$ as approximated by the dashed line in the
first plot. Similarly,  ${\cal L}_r$ tends towards the differential entropy
$H(Z)$ of the prior (where $H(Z) = \log(\sigma \sqrt{2 \pi e}) = 1.419$ in the
case of a gaussian with unit variance), as shown by the dashed line in the
second plot.
Finally, let us note that the ideal situation of a
classifier that is both optimal and pivotal appears to be unreachable for this
problem, as shown in the third plot by the  offset between ${\cal L}_f - \lambda
{\cal L}_r$ and the dashed line approximating $H({Y|X}) -
\lambda H(Z)$.

@@

In collisions at the Large Hadron Collider, massive particles can be produced at
such high energies that their decay products are collimated and the resulting
jets overlap. Still, deducing whether an observed jet is either due  to a single
hadronic particle or to the decay of a massive object into multiple hadronic
particles  is critical for understanding the nature of the particles produced in
the collision \citep{butterworth2008jet,guest2016jet}. More importantly, this
classification problem is often made difficult by the presence of simulatenous
interactions, known as pileup, which contributes significant energy depositions
unrelated to the particles of interest.

\begin{figure}
    \includegraphics[width=0.48\textwidth]{figures/ams.pdf}
    \vspace{-1cm}
    \caption{(to write)}
    \label{fig:physics-ams}
\end{figure}

>>>>

In this work, we repurpose adversarial training as a means to constraint the
predictive model $f$ in order to satisfy Eqn.~\ref{eqn:criterion}. As
illustrated in Fig.~\ref{fig:architecture}, we pit $f$ against an adversary
model $r := p_{\theta_r}(z | f(X;\theta_f)=s)$ of parameters $\theta_r$ and
associated loss ${\cal L}_r(\theta_f, \theta_r)$. This model takes  as input
realizations $s$ of $f(X; \theta_f)$, for the current value $\theta_f$ of $f$
parameters, and produces as output a function $p_{\theta_r}(z | f(X;\theta_f)=s)$
modeling the posterior probability density that $z$ parameterizes the sample
observed as $s$.
Intuitively, if $p(f(X; \theta_f)=s|z)$ varies with $z$,
then the corresponding correlation can be captured by $r$. By contrast, if
$p(f(X; \theta_f)=s|z)$ is invariant with $z$, as we require, then $r$ should
perform poorly and be close to random guessing. Training $f$ such that it
additionally minimizes the performance of $r$ therefore acts as a regularization
towards Eqn.~\ref{eqn:criterion}.

@@

On many problems of interest, let us also note that such a classifier may
not exist because the nuisance parameter directly shapes the decision boundary,
in which cases the lower bound $ H({Y|X}) - H(Z) < {\cal L}_f(\theta_f) - H({Z|f(X;\theta_f)})$ is strict: $f$ can either be an optimal classifier or a
pivotal quantity, but not both simultaneously. In this situation, it is natural
to rewrite the value function $E$  as
\begin{equation}\label{eqn:vf-lambda}
    E_\lambda(\theta_f, \theta_r) = {\cal L}_f(\theta_f) - \lambda {\cal L}_r(\theta_f, \theta_r),
\end{equation}
where $\lambda \geq 0$ is a hyper-parameter controlling the trade-off between
the performance of $f$ and its independence with respect to the nuisance
parameter. Setting $\lambda$ to a large value will preferably enforces $f$ to
be pivotal while setting $\lambda$ close to $0$ will rather constraint $f$ to be
optimal.

@@

Assuming a gaussian prior $z \sim {\cal N}(0,1)$ of unit variance, we start by
generating training data $\{ x_i, y_i, z_i \}_{i=1}^N$, from which we build a
neural network classifier $f$ minimizing ${\cal L}_f(\theta_f)$ without
adversarial training. The network architecture comprises 2 dense hidden layers of 20
nodes with ReLU activations, followed  by a dense output layer with a single
node with a sigmoid activation. As shown in the upper plots of
Fig.~\ref{fig:toy}, the resulting classifier is not pivotal, as the
conditional probability densities of its decision scores $f(X;\theta_f)$ show
large discrepancies between values $z$ of the nuisance. While not shown here, a
classifier trained only from data generated at the nominal value $Z=0$ would
also not be pivotal.

Let us now consider the adversarial training of $f$ against an adversary $r$
implemented as a mixture density network modeling $Z|f(X;\theta_f)$ as a mixture
of five gaussians. As for $f$, the network architecture comprises 2 dense hidden
layers of 20 nodes with ReLU activations, but is followed by an output layer of
15 nodes corresponding to the means, standard deviations and mixture
coefficients of the five gaussians. Output nodes for the mean values come with
linear activations, nodes for the standard deviations with exponential
activations to ensure positivity, while nodes for the mixture coefficients
implement the softmax function to ensure positivity and normalization. When
running Algorithm~\ref{alg:adversarial-training} as initialized with the
classifier $f$ obtained previously, adversarial training effectively reshapes
the decision function so it that becomes almost independent on the nuisance
parameter, as shown in the lower plots of Fig.~\ref{fig:toy}. In particular,
the conditional probability densities of the decision scores $f(X;\theta_f)$ are
now very similar to each other, indicating only a small residual  dependency on the
nuisance, as theoretically expected. The dynamics of adversarial training is
illustrated in Fig.~\ref{fig:toy-training}, where the losses ${\cal L}_f$,
${\cal L}_r$ and ${\cal L}_f - \lambda {\cal L}_r$ are evaluated after each
iteration of Algorithm~\ref{alg:adversarial-training}. In the first iterations,
we observe that the global objective ${\cal L}_f - \lambda {\cal L}_r$ is
minimized by making the classifier less accurate, hence the corresponding
increase of ${\cal L}_f$, but which results in a classifier that is more
pivotal, hence the corresponding increase of ${\cal L}_r$ and the total net
benefit. As learning goes, minimizing $E$ then requires making predictions that
are more accurate, hence decreasing ${\cal L}_f$, or that are even less dependent on
$Z$, hence shaping $p_{\theta_r}$ towards the prior $p(z)$. Indeed, ${\cal L}_f$
eventually starts to slightly decrease, while remaining lower bounded by
$\min_{\theta_f} {\cal L}_f(\theta_f)$ as approximated by the dashed line in the
first plot. Similarly,  ${\cal L}_r$ tends towards the differential entropy
$H(Z)$ of the prior (where $H(Z) = \log(\sigma \sqrt{2 \pi e}) = 1.419$ in the
case of a gaussian with unit variance), as shown by the dashed line in the
second plot.
Finally, let us note that the ideal situation of a
classifier that is both optimal and pivotal appears to be unreachable for this
problem, as shown in the third plot by the  offset between ${\cal L}_f - \lambda
{\cal L}_r$ and the dashed line approximating $H({Y|X}) -
\lambda H(Z)$.

@@

In collisions at the Large Hadron Collider, massive particles can be produced at
such high energies that their decay products are collimated and the resulting
jets overlap. Still, deducing whether an observed jet is either due  to a single
hadronic particle or to the decay of a massive object into multiple hadronic
particles  is critical for understanding the nature of the particles produced in
the collision \citep{butterworth2008jet,guest2016jet}. More importantly, this
classification problem is often made difficult by the presence of simultaneous
interactions, known as pileup, which contributes significant energy depositions
unrelated to the particles of interest.

In this setting, we reuse the classification problem of \citep{guest2016jet}
between single jets produced in quark or gluon fragmentation and two overlapping
jets produced when a high-velocity $W$ boson decays to a collimated pair of
quarks. As nuisance parameter, we consider events without pileup ($Z=0$) and
events with pileup ($Z=1$), for which an average number of $\langle \mu \rangle =
50$ unrelated interactions are overlaid. Our goal is to build an accurate
classifier, for which we also want to minimize the effects due to the
uncertainties on the nuisance. More specifically, we choose to recast the
classification problem as a hypothesis test between signal+background (jets
originating from $W$ bosons or from single quarks and gluons) and background
only (jets originating from single quarks and gluons only) and tune the decision
threshold of our classifier by maximizing its approximate median significance
(AMS), when uncertainties in the background are taken into account (see Eqn.~19
of \citep{adam2014higgs}). Our motivation is that reducing the effects of
uncertainties by requiring independence of $Z$ with the classifier output should
allow for a larger maximum significance.

To minimize the effects of the nuisance $Z$ in the background events, we train a
classifier using  Algorithm~\ref{alg:adversarial-training} but consider the
adversarial term ${\cal L}_r$ conditioned on $Y=0$ only. Both $f$ and $r$ are
neural networks with 3 hidden layers of 64 nodes and ReLU activations, each
terminated by a single final output node with a sigmoid activation. Training is
performed on a subset of 150000 samples while AMS is evaluated on an independent
test set of 5000000 samples. Results reported below are averages over 5 runs.

As shown in  Fig.~\ref{fig:physics-ams}, trading classification accuracy for
independence with respect to pileup results in a positive total net benefit in
terms of statistical significance. Without adversarial training (at
$\lambda=0$), the maximum significance peaks at $7$. By contrast, as the
independence constraint is made stronger (for $\lambda > 0$) the AMS peak moves
higher, with a maximum value around  $7.8$ for $\lambda=10$. In other words,
sacrificing accuracy to reduce  the  effects of the systematics in the
background jets does indeed allow one to improve the overall significance of the
results. Setting $\lambda$ too high however (e.g. $\lambda=500$) results in a
decrease of the maximum significance, by focusing $f$ too much on independence,
at the expense of too large a drop in accuracy.

\begin{figure}
    \includegraphics[width=0.48\textwidth]{figures/ams.pdf}
    \vspace{-1cm}
    \caption{Approximate median signifance as a function of the decision threshold
             on the output of $f$.  As shown at $\lambda=10$, trading
             classification accuracy for independence with respect to pileup
             results in a positive total net benefit in terms of statistical significance.}
    \label{fig:physics-ams}
\end{figure}
